Why do species get a thin slice of $\pi$? Revisiting Lewontin's Paradox of Variation

### 2021-02-16 manuscript/manuscript.tex
minor corrections to ms, thanks matt!

\begin{abstract}

  Under neutral theory, the level of polymorphism in an equilibrium population
  is expected to increase with population size. However, observed levels of
  diversity across metazoans vary only two orders of magnitude, while census
  population sizes ($N_c$) are expected to vary over several. This unexpectedly
  narrow range of diversity is a longstanding enigma in evolutionary genetics
  known as Lewontin's Paradox of Variation (1974). Since Lewontin's
  observation, it has been argued that selection constrains diversity across
  species, yet tests of this hypothesis seem to fall short of explaining the
  orders-of-magnitude reduction in diversity observed in nature. In this work,
  I revisit Lewontin's Paradox and assess whether current models of linked
  selection are likely to constrain diversity to this extent.  To quantify the
  discrepancy between pairwise diversity and census population sizes across
  species, I combine genetic data from 172 metazoan taxa with estimates of
  census sizes from geographic occurrence data and population densities
  estimated from body mass.  Next, I fit the relationship between
  previously-published estimates of genomic diversity and these approximate
  census sizes to quantify Lewontin's Paradox. While previous across-taxa
  population genetic studies have avoided accounting for phylogenetic
  non-independence, I use phylogenetic comparative methods to investigate the
  diversity census size relationship, estimate phylogenetic signal, and explore
  how diversity changes along the phylogeny. I consider whether the reduction
  in diversity predicted by models of recurrent hitchhiking and background
  selection could explain the observed pattern of diversity across species.
  Since the impact of linked selection is mediated by recombination map length,
  I also investigate how map lengths vary with census sizes. I find species
  with large census sizes have shorter map lengths, leading these species to
  experience greater reductions in diversity due to linked selection. Even
  after using high estimates of the strength of sweeps and background
  selection, I find linked selection likely cannot explain the shortfall
  between predicted and observed diversity levels across metazoan species.
  Furthermore, the predicted diversity under linked selection does not fit the
  observed diversity--census-size relationship, implying that processes other
  than background selection and recurrent hitchhiking must be limiting
  diversity.

@@

A resolution of Lewontin's Paradox would involve a mechanistic description and
quantification of the evolutionary processes that prevent diversity from
scaling with census sizes across species. This would necessarily connect to the
broader literature on the empirical relationship between diversity and
population size \parencite{Frankham1996-yb,Nei1984-zi,Soule1976-he}, and the
ecological and life history correlates of genetic diversity
\parencite{Nevo1978-wh,Powell1975-lg,Nevo1984-hp}. Three categories of
processes stand out as potentially capable of decoupling census sizes from
diversity: non-equilibrium demography, variance and skew in reproductive
success, and selective processes.

@@

% Furthermore, Coop's re-analysis finds no evidence
% of a remaining correlation between recombination and diversity among the
% highest recombination windows, as would be expected if these $\pi_0$ estimates
% were still impacted by linked selection.

@@

Finally, I explore whether the predicted reductions of diversity under
background selection and recurrent hitchhiking are sufficiently strong enough
to resolve Lewontin's Paradox. These predicted reductions in diversity across
species are generously estimated using selection strength parameters from
\emph{Drosophila melanogaster}, a species known to be strongly affected by
linked selection. Given the effects of linked selection are mediated by
recombination map length, I investigate how recombination map lengths vary with
census population size using data from a previously-published survey
\parencite{Stapley2017-fs}. I find map lengths are typically shorter in
large--census-size species, increasing the effects of linked selection in these
species, which might further decouple diversity from census size. Still, I find
the combined impact of these selection models with available parameter
estimates falls short in explaining Lewontin's Paradox, and discuss future
avenues through which the Paradox of Variation could be fully resolved.

@@

To determine which ecological or evolutionary processes could decouple
diversity from census population size, we first need to quantify this
relationship across a wide variety of taxa. Previous work has found there is a
significant relationship between heterozygosity and the logarithm of population
size, but these studies relied on heterozygosity measured from allozyme data
\parencite{Soule1976-he,Frankham1996-yb,Nei1984-zi}. Here, I confirm these
findings using pairwise diversity estimates from genomic sequence data and the
estimated census sizes (Figure \ref{fig:figure-1}). The pairwise diversity
estimates are from three sources: \textcite{Leffler2012-zj},
\textcite{Corbett-Detig2015-gt}, and \textcite{Romiguier2014-bp}, and are
predominantly from either synonymous or non-coding DNA (see Methods:
\Newnameref{sec:methods-diversity}). Overall, an ordinary least squares (OLS)
relationship on a log-log scale fits the data well (Figure \ref{fig:figure-1}).
The OLS slope estimate is significant and implies an increase in diversity of
0.09 differences per basepair for every order of magnitude census size grows
(95\% confidence interval $[0.08, 0.12]$; see also the OLS fit per-phyla,
Supplementary Figure \ref{suppfig:figure-1-ave}).

@@

Using a phylogenetic mixed-effects model
\parencite{Lynch1991-tp,Hadfield2010-ql,De_Villemereuil2014-kt} implemented in
Stan \parencite{Carpenter2017-xi,Stan_Development_Team2020-ea}, I estimated the
linear relationship between diversity and population size (on a log-log scale)
accounting for phylogeny, for the 166 taxa with non-missing data and present in
the synthetic chronogram. As with the non-phylogenetic regression, this
relationship was positive and significant (95\% credible interval $0.04,
0.11$), though somewhat attenuated compared to the OLS estimates (Figure
\ref{fig:figure-1}B). Since the population size estimates are based on range
and body mass, they are essentially a composite trait; fitting phylogenetic
mixed-effects models separately on body mass and range indicates these have
significant negative and positive effects, respectively (Supplementary Figure
\ref{suppfig:figure-div-range-bodymass}).

@@

One limitation of the phylogenetic mixed-effects models employed here is that
they assume traits evolve under constant-rate Brownian motion. To test this
assumption, I performed node-height tests \parencite{Freckleton2006-ww}.
Node-height tests regress the absolute values of the standardized contrasts
between lineages against the branching time (since present) of these lineages.
Under Brownian Motion (BM), standardized contrasts are estimates of the rate of
character evolution \parencite{Felsenstein1985-an}; if a trait evolves under
constant rate BM, this relationship should be flat. For both diversity and
population size, node-height tests indicate a significant increase in the rate
of evolution towards the present (robust regression p-values 0.028 and 0.00070
respectively; Figure \ref{fig:figure-2}C).  Considering the constituents of the
population size estimate, range and body mass, separately, range but not body
mass shows a significant increase (p-value $1.9 \times 10^{-7}$) in rate
towards the present.

@@

I quantify the effect of linked selection on diversity as the ratio of observed
diversity ($\pi$) to the estimated diversity in the absence of linked selection
($\pi_0$), $R = \nicefrac{\pi}{\pi_0}$. There are two difficulties in
evaluating whether linked selection could resolve Lewontin's Paradox. The first
difficulty is that $\pi_0$ is unobserved. Previous work has estimated $\pi_0$
using methods that exploit the spatial heterogeneity in recombination and
functional density across the genome to fit linked selection models that
incorporate both hitchhiking and background selection
\parencite{Elyashiv2016-vt,Corbett-Detig2015-gt}. The second difficulty is
understanding how $R$ varies across taxa, since we lack estimates of critical
model parameters for most species. Still, I can address a key question: if
diversity levels were determined by census sizes ($\pi_0 = 4N_c \mu$), are the
combined effects of background selection and recurrent hitchhiking sufficient
to reduce diversity to observed levels? Furthermore, does the relationship
between census size and predicted diversity under linked selection across
species, $\pi_{BGS+HH} = R \pi_0$, match the observed relationship in Figure
\ref{fig:figure-1}?

@@

Parameterizing the model this way, I then set the key parameters that
determine the impact of recurrent hitchhiking and background selection
($\gamma$, $J$, and $U$) to high values estimated from \emph{Drosophila
melanogaster} by \textcite{Elyashiv2016-vt}. My estimate of
$\gamma_\text{Dmel}$ based Elyashiv et al. implies $\nu_\text{BP,Dmel} \approx
2.34 \times 10^{-11}$, which is close to other estimates from \emph{D.
melanogaster} (see Supplementary Figure \ref{suppfig:linked-sel-params}A). The
rate of deleterious mutations per diploid genome, per generation is
parameterized using the estimate from Elyashiv et al., $U_\text{Dmel} = 1.6$,
which is a bit greater than previous estimates based on Bateman-Mukai
approaches \parencite{Mukai1985-bc,Mukai1988-vs,Charlesworth1987-ab}. Finally,
the probability that a lineage is trapped in a sweep, $J_\text{Dmel}$, is
calculated from the estimated genome-wide average coalescent rate due to sweeps
from Elyashiv et al. (see Supplementary Figured
\ref{suppfig:linked-sel-params}B and Methods:
\Newnameref{sec:methods-reduction} for more details on parameter estimates).
Using these \emph{Drosophila} parameters, I then explore how the predicted
range of diversity levels under background selection and recurrent hitchhiking
varies across species with recombination map length ($L$) and census population
size ($N_c$).

@@

Lewontin's Paradox shows the extent to which the effective population sizes
implied by diversity, $\widetilde{N}_e$, diverge from census population sizes.
However, there are a variety other effective population size estimates
calculable from different data and summary statistics
\parencite{Wang2016-mi,Caballero1994-ao,Caballero2020-wm}. These include
estimators based on the observed decay in linkage disequilibrium or temporal
estimators that use the variance in allele frequency change. These alternate
estimators capture summaries of the effective population size on shorter
timescales than coalescent-based estimators \parencite{Wang2005-wy}, and thus
could be used to tease out processes that impact the $N_e$-$N_c$ relationship
in the more recent past.

@@

I would like to thank Andy Kern and Peter Ralph for helpful discussions and
supporting me during this work, and Graham Coop for inspiration and helpful
feedback during socially distanced nature walks at Yolo Basin. I thank Jessica
Stapley for kindly providing the recombination map length data, and Yaniv
Brandvain, Amy Collins, Doc Edge, Tyler Kent, Chuck Langley, Sally Otto, Jeff
Ross-Ibarra, Aaron Stern, Anastasia Teterina, Michael Turelli, and my
Kern-Ralph labmates for helpful discussions. Sarah Friedman, Katherine Corn,
and Josef Uyeda provided very useful advice about phylogenetic comparative
methods; yet I take full responsibility for any shortcomings of my analysis.  I
would like to also thank UO librarian Dean Walton for helping me track down
some rather difficult to find older papers. This work was supported by an NIH
Grant (1R01GM117241) awarded to Andrew Kern.

>>>>

% Neutral theory predicts that genetic diversity increases with population size,
% yet observed levels of diversity across species vary only two orders of
% magnitude while population sizes likely vary over several. The causes of this
% discrepancy, known as Lewontin's Paradox of Variation, remain unknown. Here I
% revisit Lewontin's Paradox by quantifying the relationship between pairwise
% diversity and approximate census size for 172 metazoan species. Using
% phylogenetic comparative methods, I show this relationship is significant
% accounting for phylogeny, but has high phylogenetic signal and some lineages
% experience shifts in diversity deep in the past. I find a negative relationship
% between recombination map length and census size, suggesting abundant species
% have less recombination and could experience greater reductions in diversity
% due to linked selection. However, I show that even using strong selection
% parameter estimates, models of linked selection are unlikely to explain the
% observed relationship between diversity and census sizes across species.

\begin{abstract} 
  Under neutral theory, genetic diversity in an equilibrium population
  is expected to increase with population size. However, observed levels of
  diversity across metazoans vary only two orders of magnitude, while census
  population sizes ($N_c$) are expected to vary over several. This unexpectedly
  narrow range of diversity is a longstanding enigma in evolutionary genetics
  known as Lewontin's Paradox of Variation (1974). Since Lewontin's
  observation, it has been argued that selection constrains diversity across
  species, yet tests of this hypothesis seem to fall short of explaining the
  orders-of-magnitude reduction in diversity observed in nature. In this work,
  I revisit Lewontin's Paradox and assess whether current models of linked
  selection are likely to constrain diversity to this extent.  To quantify the
  discrepancy between pairwise diversity and census population sizes across
  species, I combine genetic data from 172 metazoan taxa with estimates of
  census sizes from geographic occurrence data and population densities
  estimated from body mass.  Next, I fit the relationship between
  previously-published estimates of genomic diversity and these approximate
  census sizes to quantify Lewontin's Paradox. While previous across-taxa
  population genetic studies have avoided accounting for phylogenetic
  non-independence, I use phylogenetic comparative methods to investigate the
  diversity census size relationship, estimate phylogenetic signal, and explore
  how diversity changes along the phylogeny. I consider whether the reduction
  in diversity predicted by models of recurrent hitchhiking and background
  selection could explain the observed pattern of diversity across species.
  Since the impact of linked selection is mediated by recombination map length,
  I also investigate how map lengths vary with census sizes. I find species
  with large census sizes have shorter map lengths, leading these species to
  experience greater reductions in diversity due to linked selection. Even
  after using high estimates of the strength of sweeps and background
  selection, I find linked selection likely cannot explain the shortfall
  between predicted and observed diversity levels across metazoan species.
  Furthermore, the predicted diversity under linked selection does not fit the
  observed diversity--census-size relationship, implying that processes other
  than background selection and recurrent hitchhiking must be limiting
  diversity.

@@

A resolution of Lewontin's Paradox would involve a mechanistic description and
quantification of the evolutionary processes that prevent diversity from
scaling with census sizes across species. This would necessarily connect to the
broader literature on the empirical relationship between diversity and
population size
\parencite{Frankham1996-yb,Nei1984-zi,Soule1976-he,Leroy2021-gy}, and the
ecological and life history correlates of genetic diversity
\parencite{Nevo1978-wh,Powell1975-lg,Nevo1984-hp}. Three categories of
processes stand out as potentially capable of decoupling census sizes from
diversity: non-equilibrium demography, variance and skew in reproductive
success, and selective processes.

@@

@@

Finally, I explore whether the predicted reductions of diversity under
background selection and recurrent hitchhiking are sufficiently strong 
to resolve Lewontin's Paradox. These predicted reductions in diversity across
species are generously estimated using selection strength parameters from
\emph{Drosophila melanogaster}, a species known to be strongly affected by
linked selection. Given the effects of linked selection are mediated by
recombination map length, I investigate how recombination map lengths vary with
census population size using data from a previously-published survey
\parencite{Stapley2017-fs}. I find map lengths are typically shorter in
large--census-size species, increasing the effects of linked selection in these
species, which might further decouple diversity from census size. Still, I find
the combined impact of these selection models with available parameter
estimates falls short in explaining Lewontin's Paradox, and discuss future
avenues through which the Paradox of Variation could be fully resolved.

@@

To determine which ecological or evolutionary processes could decouple
diversity from census population size, we first need to quantify this
relationship across a wide variety of taxa. Previous work has found there is a
significant relationship between heterozygosity and the logarithm of population
size, but these studies relied on heterozygosity measured from allozyme data
\parencite{Soule1976-he,Frankham1996-yb,Nei1984-zi}. Here, I confirm these
findings using pairwise diversity estimates from genomic sequence data and the
estimated census sizes (Figure \ref{fig:figure-1}). The pairwise diversity
estimates are from three sources: \textcite{Leffler2012-zj},
\textcite{Corbett-Detig2015-gt}, and \textcite{Romiguier2014-bp}, and are
predominantly from either synonymous or non-coding DNA (see Methods:
\Newnameref{sec:methods-diversity}). Overall, an ordinary least squares (OLS)
relationship on a log-log scale fits the data well (Figure \ref{fig:figure-1}).
The OLS slope estimate is significant and implies a 0.09 percent increase in
differences per basepair for every order of magnitude census size grows (95\%
confidence interval $[0.08, 0.12]$; see also the OLS fit per-phyla,
Supplementary Figure \ref{suppfig:figure-1-ave}).

@@

Using a phylogenetic mixed-effects model
\parencite{Lynch1991-tp,Hadfield2010-ql,De_Villemereuil2014-kt} implemented in
Stan \parencite{Carpenter2017-xi,Stan_Development_Team2020-ea}, I estimated the
linear relationship between diversity and population size (on a log-log scale)
accounting for phylogeny, for the 166 taxa with non-missing data and present in
the synthetic chronogram. As with the non-phylogenetic regression, this
relationship was positive and significant (95\% credible interval $0.04,
0.11$), though somewhat attenuated compared to the OLS estimates (Figure
\ref{fig:figure-2}B). Since the population size estimates are based on range
and body mass, they are essentially a composite trait; fitting phylogenetic
mixed-effects models separately on body mass and range indicates these have
significant negative and positive effects, respectively (Supplementary Figure
\ref{suppfig:figure-div-range-bodymass}).

@@

One limitation of the phylogenetic mixed-effects models employed here is that
they assume traits evolve under constant-rate Brownian motion. To test this
assumption, I performed node-height tests \parencite{Freckleton2006-ww}.
Node-height tests regress the absolute values of the standardized contrasts
between lineages against the branching time (since present) of these lineages.
Under Brownian Motion (BM), standardized contrasts are estimates of the rate of
character evolution \parencite{Felsenstein1985-an}; if a trait evolves under
constant rate BM, this relationship should be flat. For both diversity and
population size, node-height tests indicate a significant increase in the rate
of evolution towards the present (robust regression p-values 0.028 and 0.00070
respectively; Figure \ref{fig:figure-2}C).  Considering the constituents of the
population size estimate, range and body mass, separately, the rate of
evolution of range but not body mass shows a significant increase (p-value $1.9
\times 10^{-7}$) towards the present.

@@

I quantify the effect of linked selection on diversity as the ratio of observed
diversity ($\pi$) to the estimated diversity in the absence of linked selection
($\pi_0$), $R = \nicefrac{\pi}{\pi_0}$. There are two difficulties in
evaluating whether linked selection could resolve Lewontin's Paradox. The first
difficulty is that $\pi_0$ is unobserved. Previous work has estimated $\pi_0$
using methods that exploit the spatial heterogeneity in recombination and
functional density across the genome to fit linked selection models that
incorporate both hitchhiking and background selection
\parencite{Elyashiv2016-vt,Corbett-Detig2015-gt}. The second difficulty is
understanding of how $R$ varies across taxa, since we lack estimates of
critical model parameters for most species. Still, I can address a key
question: if diversity levels were determined by census sizes ($\pi_0 = 4N_c
\mu$), are the combined effects of background selection and recurrent
hitchhiking sufficient to reduce diversity to observed levels? Furthermore,
does the relationship between census size and predicted diversity under linked
selection across species, $\pi_{BGS+HH} = R \pi_0$, match the observed
relationship in Figure \ref{fig:figure-1}?

@@

Parameterizing the model this way, I then set the key parameters that determine
the impact of recurrent hitchhiking and background selection ($\gamma$, $J$,
and $U$) to high values estimated from \emph{Drosophila melanogaster} by
\textcite{Elyashiv2016-vt}. My estimate of $\gamma_\text{Dmel}$ based Elyashiv
et al. implies a rate of sweeps per basepair of $\nu_\text{BP,Dmel} \approx
2.34 \times 10^{-11}$, which is close to other estimates from \emph{D.
melanogaster} (see Supplementary Figure \ref{suppfig:linked-sel-params}A). The
rate of deleterious mutations per diploid genome, per generation is
parameterized using the estimate from Elyashiv et al., $U_\text{Dmel} = 1.6$,
which is a bit greater than previous estimates based on Bateman-Mukai
approaches \parencite{Mukai1985-bc,Mukai1988-vs,Charlesworth1987-ab}. Finally,
the probability that a lineage is trapped in a sweep, $J_\text{Dmel}$, is
calculated from the estimated genome-wide average coalescent rate due to sweeps
from Elyashiv et al. (see Supplementary Figured
\ref{suppfig:linked-sel-params}B and Methods:
\Newnameref{sec:methods-reduction} for more details on parameter estimates).
Using these \emph{Drosophila} parameters, I then explore how the predicted
range of diversity levels under background selection and recurrent hitchhiking
varies across species with recombination map length ($L$) and census population
size ($N_c$).

@@

Lewontin's Paradox shows the extent to which the effective population sizes
implied by diversity, $\widetilde{N}_e$, diverge from census population sizes.
However, there are a variety other effective population size estimates
calculable from different data and summary statistics
\parencite{Wang2016-mi,Caballero1994-ao,Caballero2020-wm,Galtier2020-fb}. These
include estimators based on the observed decay in linkage disequilibrium or
temporal estimators that use the variance in allele frequency change. These
alternate estimators capture summaries of the effective population size on
shorter timescales than coalescent-based estimators \parencite{Wang2005-wy},
and thus could be used to tease out processes that impact the $N_e$-$N_c$
relationship in the more recent past.

@@

I would like to thank Andy Kern and Peter Ralph for helpful discussions and
supporting me during this work, and Graham Coop for inspiration and helpful
feedback during socially distanced nature walks at Yolo Basin. I thank Jessica
Stapley for kindly providing the recombination map length data, and Yaniv
Brandvain, Amy Collins, Doc Edge, Tyler Kent, Chuck Langley, Matt Osmond, Sally
Otto, Jeff Ross-Ibarra, Aaron Stern, Anastasia Teterina, Michael Turelli, and
my Kern-Ralph labmates for helpful discussions. Sarah Friedman, Katherine Corn,
and Josef Uyeda provided very useful advice about phylogenetic comparative
methods; yet I take full responsibility for any shortcomings of my analysis.  I
would like to also thank UO librarian Dean Walton for helping me track down
some rather difficult to find older papers. This work was supported by an NIH
Grant (1R01GM117241) awarded to Andrew Kern.
